---
title: "Introduction to MDAW"
output: rmarkdown::html_vignette
vignette: >
    %\VignetteIndexEntry{Introduction to MDAW}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

In metabolomics studies, the accurate analysis and interpretation
of complex data sets is of utmost importance in order to obtain meaningful 
biological insights. To address this issue, a robust computational workflow 
for statistical analysis of both untargeted and targeted metabolomics data 
is essential. Such a workflow not only ensures the reliability of results, but 
also promotes reproducibility in data analysis and reporting - `a fundamental
principle for scientific integrity` (Resnik DB. et al., 2016). 

In this context, the development of a comprehensive computational framework 
is necessary, This framework should seamlessly integrate various analysis 
steps, from data preprocessing to advanced statistical modeling, while adhering 
to best practices in data processing and interpretation. This allows the users 
to clearly and confidently navigate the intricacies of metabolomics data, 
enabling in-depth scientific investigation and knowledge expansion. 

In this manuscript, a carefully designed computational workflow was developed
to the analysis of metabolomics data in binary classification studies. 
The approach encompasses a range of techniques, including **univariate** 
and **multivariate** statistical analyses, **quality control** procedures, 
and **visualization** tools, all aimed at deciphering the complex metabolic 
signatures underlying biological phenomena. By providing a structured 
framework and leveraging state-of-the-art methods enhances the robustness 
of data analysis in metabolomics.

Here are the outlines for `MDAW`, a workflow for analyzing metabolomics data, 
in a binary classification study. 
Let's see together the breakdown steps which consists this workflow:

1. **Importing Data**: Start by importing both metabolite and experimental data
from an Excel sheet. This step ensures you have all needed information 
for the analysis.

2. **Pooled QC-based Data Cleaning**: Quality control (QC) is important 
in metabolomics to make sure that data obtained are reliable. 
Pooled QC samples helps to identify and remove any technical variation 
or artifacts in the data.

3. **Principal Component Analysis (PCA) Visualization**: PCA is a powerful 
tool for `assessing data quality`, `identifying outliers`, 
and `gaining insights into the overall structure of the data`. 
Visualizing the data using PCA plots help us to detect any clustering 
or separation between sample groups.

4. **Two-Class Univariate Statistics**: Perform uni-variate statistical 
analysis to identify individual metabolites that show significant differences
between the two classes (e.g., control vs. treatment). This analysis can
involve statistical tests such as `t-tests` or `ANOVA`, followed 
by correction for multiple comparisons if needed.

5. **Multivariate Analysis with PLS-DA**:

    - **Model Optimization (R2 vs. Q2)**: Optimize the PLS-DA model by assessing
    the goodness of fit (R2) and predictive ability (Q2) using cross-validation
    techniques.

    - **Permutation Testing**: Validate the PLS-DA model by permutation testing
    to assess its statistical significance and guard against overfitting.

    - **Model Prediction Metrics**: Evaluate the performance of the PLS-DA model
    using metrics such as accuracy, sensitivity, specificity, and area under
    the receiver operating characteristic curve (AUC-ROC).

    - **Feature Importance**: Identify important metabolites contributing
    to the classification by examining variable importance in projection (VIP)
    scores or loading plots.

    - **Model Prediction Data Visualizations**: Visualize the results
    of the PLS-DA model, such as score plots and class prediction plots,
    to assess the separation between sample groups
    and the model's predictive ability.

6. **Exporting Statistical Tables**: Export relevant statistical tables
summarizing the results of univariate and multivariate analyses
to Excel sheets for further examination and reporting.

This workflow should provide a well streamlined computational approach
to analyzing metabolomics data in a binary classification study,
integrating both univariate and multivariate statistical methods
to identify biomarkers and gain insights into biological differences
between sample groups.

<details>
<summary>**Session Info**</summary>
```{r sessioninfo}
library(sessioninfo)
sessioninfo::session_info()
```
</details>

## Libraries

Libraries(or Packages) provide functions that extend the basic functionality 
of the R language. We will need the following tools to analyse the data:


```{r}
library(MDAW)
library(dplyr)
```


## Import data/peak

The chunk code below utilises a function called `import_dataXL()` to import
the Peak and Data sheets from an Excel file. Once this is done, a notification
appears to confirm that the Peak and Data tables have been loaded
from the corresponding worksheets in the Excel file.

```{r}
# Theinput file (Excel spreadsheet)
excel_file <- "Gastric_NMR.xlsx"

# The path to the input file (Excel spreadsheet)
path <- system.file("extdata", "Gastric_NMR.xlsx", package = "MDAW")

# The table of Data from an Excel file(data sheet)
table_data <- MDAW::import_dataXL(xlsx_file_name = path, sheet_name = "Data", 
                                  interactive_output = FALSE)

# The table of Peak from an Excel file(peak sheet)
table_peak <- MDAW::import_dataXL(xlsx_file_name = path, sheet_name = "Peak")
```

## Load table of data

The data table offers an interactive display feature, allowing to easily 
inspect and verify the imported values. This functionality can be accessed
by utilizing the `reactable()` function.

```{r}
table_data  # View and check the table data 
```


## Load table of peaks

The peak table offers an interactive display feature, allowing to easily 
inspect and verify the imported values.

```{r}
table_peak  # View and check the table data 
```

## Data Cleansing

Before beginning statistical or machine learning modeling, it is important to assess the quality of your data and eliminate any inaccurately measured metabolites. This process ensures the integrity of your analytics. For this demonstration using the gastric cancer NMR data set, basic statistical to measure each metabolite are already calculated, which were then stored in the `table_peak.` In this file, metabolites are stored in compliance with the following conditions:

    - A QC-RSD (Relative Standard Deviation) below 20%.
    
    - Fewer than 10% of values are absent.
    
```{r}
# filter clean metabolite from table_peak
clean_table_peak <-
  MDAW::clean_table_peak(table = table_peak,
                         min_QC_RSD = 20,
                         min_Perc_missing = 20)
```

## PCA (Principal Component Analysis)

```{r}
list_of_peaks <- clean_table_peak["Name"]
table_data |>
    dplyr::select(RowID, 
                  Idx, 
                  SampleID, 
                  SampleType, 
                  Class,
                  list_of_peaks$Name)

matrix_data <- table_data |>
    dplyr::select(list_of_peaks$Name)

log10(matrix_data)

# missing value imputation (knn - 3 nearest neighbors)
```

